---
title: "Hypothesis_Testing"
author: "Tanishq"
date: "2025-05-02"
output: html_document
---

```{r}
library(tidyverse)
library(infer)
sat <- read_csv("SAT.csv")
head(sat)
```

# Data Cleaning

```{r}
sat <- sat |> rename(reading = `Critical Reading Mean`,
                     math = `Mathematics Mean`,
                     writing = `Writing Mean`) |>
  filter(!is.na(reading), !is.na(math), !is.na(writing))
  
sat_aov <- sat |> select(reading, 
                         math, 
                         writing) |>
  pivot_longer(cols = c(reading, 
                        math, 
                        writing),
               names_to = "subject",
               values_to = "score")
  
  head(sat_aov)

sat_math_reading <- sat |> 
  select(math, reading) |> 
  mutate(diff = math - reading)

sat_math_writing <- sat |> 
  select(math, writing) |> 
  mutate(diff = math - writing)

sat_writing_reading <- sat |> 
  select(writing, reading) |> 
  mutate(diff = writing - reading)
```

# Anova Test

Significance level $\alpha = 0.05$ for all the tests.

$H_{0}:$ The mean SAT score is same across all the subject.

$\mu_{math} = \mu_{reading} = \mu_{writing}$

$H_{a}:$ At least one of the mean SAT score is different from others.

At least one $\mu$ differs among math, reading and writing

```{r}
anova_result <- aov(formula = score~subject, data = sat_aov)
summary(anova_result)
```

# Anova Conclusion

since p-value $= 0.00194 < 0.05$ We Reject Null Hypothesis.

At least one subject's mean SAT score (Math, Reading, Writing) is significantly different from others.

# Pairwise T-Test

why? to find out which subject's mean SAT score differs from other.

```{r}
pairwise.t.test(sat_aov$score, 
                sat_aov$subject, 
                p.adjust.method = "bonferroni")
```

# Pairwise T-Test Conclusion

| Comparison         | Adjusted p-value | Result                    |
|--------------------|------------------|---------------------------|
| Math vs Reading    | 0.1332           | No Significant Difference |
| Math vs Writing    | 0.0013           | Significant Difference    |
| Writing vs Reading | 0.3866           | No Significant Difference |

Only Math and Writing shows significant difference in mean SAT score.

# Box Plot

To visualize difference between mean SAT scores across each subject.

```{r}
ggplot(sat_aov, aes(x = subject, y = score, fill = subject)) +
  geom_boxplot() +
  labs(title = "SAT Scores by Subject",
       x = "Subject",
       y = "SAT Score") +
  theme(legend.position = "none")
```

# Paired T-Test

## Math vs Writing

$H_{0}$: There is no difference in Mathematics mean score and Writing Mean score across all schools

$H_{a}$: There is significant difference in Mathematics mean score and Writing Mean score across all schools

$H_{0}:$ diff $= 0$

$H_{a}:$ diff $\not= 0$

### Method 1: Using t.test (Assumes CLT condition are satisfied)

```{r}
t.test(x = sat_math_writing$math, 
       y = sat_math_writing$writing, 
       alternative = "two.sided", 
       paired = TRUE)
```

### Method 2: Using Bootstrap re-samples

```{r}
null_dist <- sat_math_writing |>
  specify(response = diff) |>
  hypothesize(null = "point", mu = 0) |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "mean")

obs_stat <- mean(sat_math_writing$diff)

get_p_value(null_dist, obs_stat, direction = "two.sided")
visualize(null_dist, bins = 20) +
  shade_p_value(obs_stat = obs_stat, direction = "two.sided")
```

## Conclusion (Math vs Writing)

The both the method above show that p value is less than the significance level therefore, rejecting the null hypothesis and concluding that difference between math and writing mean score across school is significant

## Math vs Reading

$H_{0}$: There is no difference in Mathematics mean score and Reading Mean score across all schools

$H_{a}$: There is significant difference in Mathematics mean score and Reading Mean score across all schools

$H_{0}:$ diff $= 0$

$H_{a}:$ diff $\not= 0$

### Method 1: Using t.test (Assumes CLT condition are satisfied)

```{r}
t.test(x = sat_math_reading$math, 
       y = sat_math_reading$reading, 
       alternative = "two.sided", 
       paired = TRUE)
```

### Method 2: Using Bootstrap re-samples

```{r}
null_dist <- sat_math_reading |>
  specify(response = diff) |>
  hypothesize(null = "point", mu = 0) |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "mean")

obs_stat <- mean(sat_math_reading$diff)

get_p_value(null_dist, obs_stat, direction = "two.sided")
visualize(null_dist, bins = 20) +
  shade_p_value(obs_stat = obs_stat, direction = "two.sided")
```

## Conclusion (Math vs Reading)

The both the method above show that p value is less than the significance level therefore, rejecting the null hypothesis and concluding that difference between math and reading mean score across school is significant

## Writing vs Reading

$H_{0}$: There is no difference in Writing mean score and Reading Mean score across all schools

$H_{a}$: There is significant difference in Writing mean score and Reading Mean score across all schools

$H_{0}:$ diff $= 0$

$H_{a}:$ diff $\not= 0$

### Method 1: Using t.test (Assumes CLT condition are satisfied)

```{r}
t.test(x = sat_writing_reading$writing, 
       y = sat_writing_reading$reading, 
       alternative = "two.sided", 
       paired = TRUE)
```

### Method 2: Using Bootstrap re-samples

```{r}
null_dist <- sat_writing_reading |>
  specify(response = diff) |>
  hypothesize(null = "point", mu = 0) |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "mean")

obs_stat <- mean(sat_writing_reading$diff)

get_p_value(null_dist, obs_stat, direction = "two.sided")
visualize(null_dist, bins = 20) +
  shade_p_value(obs_stat = obs_stat, direction = "two.sided")
```

## Conclusion (Writing vs Reading)

The both the method above show that p value is less than the significance level therefore, rejecting the null hypothesis and concluding that difference between writing and reading mean score across school is significant.

# Independent T-Test (Welch Two Sample T-Test)

## Math vs Writing

$H_0: \mu_{math} - \mu_{writing} = 0$

$H_a: \mu_{math} - \mu_{writing} \not= 0$

```{r}
t.test(x = sat_math_writing$math, 
       y = sat_math_writing$writing, 
       alternative = "two.sided", 
       paired = FALSE)
```

## Conclusion (Math vs Writing)

p-values $= 0.006025 < 0.05$. Reject Null Hypothesis. Mean SAT score in math and writing have significant difference.

## Math vs Reading

$H_0: \mu_{math} - \mu_{reading} = 0$

$H_a: \mu_{math} - \mu_{reading} \not= 0$

```{r}
t.test(x = sat_math_reading$math, 
       y = sat_math_reading$reading, 
       alternative = "two.sided", 
       paired = FALSE)
```

## Conclusion (Math vs Reading)

p-values $= 0.0484 < 0.05$. Reject Null Hypothesis. Mean SAT score in math and reading have significant difference.

Note: This result is different than pairwise t-test performed before why?

The result differs due to usage of Bonferroni correction made in pairwise t-test to reduce false positives.

current p_value $= 0.0484$

adjusted p_value $=$ current p_value $\cdot 3$ $= 0.1452$ which is closer to p_value in pairwise t-test for math vs reading, which was $0.1332$

## Writing vs Reading

$H_0: \mu_{writing} - \mu_{reading} = 0$

$H_a: \mu_{writing} - \mu_{reading} \not= 0$

```{r}
t.test(x = sat_writing_reading$writing, 
       y = sat_writing_reading$reading, 
       alternative = "two.sided", 
       paired = FALSE)
```

## Conclusion (Writing vs Reading)

p-values $= 0.1121 > 0.05$. Accept Null Hypothesis. Mean SAT score in writing and reading does not have significant difference.
